Ecosystem size and complexity dictate riverine biodiversity

### 2021-01-09 manuscript.Rmd
Update manuscript.Rmd

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro]
- Seoghyn Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of Minnesota]
- Taku Kadoya^[National Institute for Environmental Studies]
- Yusuke Miyazaki^[Shiraume Gakuen University]
- Jacques C. Finlay\footnotemark[2]
output:
  html_notebook: default
  word_document: default
abstract: |
  |This is the abstract.
bibliography: reference.bib
csl: nature.csl
---

@@

@@

Influences of ecosystem size and complexity differed significantly in their dependence on landscape scenarios. Ecosystem size had positive effects on $\gamma$ diversity regardless of landscape scenarios, although the slopes were steeper with greater environmental variation (higher $\sigma_h$ and/or $\sigma_l$) (). This result is attributable to the fact that larger ecosystems can hold more individuals in a metacommunity[@hubbellUnifiedNeutralTheory2001]. In contrast, we observed limited or no influences of branching complexity when local environmental noise was equal to or exceeded environmental variation at headwaters ($\sigma_l \ge \sigma_h$). Under this scenario, having more branches may not contribute to the overall habitat diversity in the ecosystem because local environmental noise masks environmental differences between branches. Therefore, this landscape scenario decouples the intimate relationship between branching structure and habitat diversity, thereby eliminating the positive effect of branching complexity on $\gamma$ diversity. This theoretical prediction may not apply to pristine or semi-natural river networks where individual streams with diverse geological and/or climatic backgrounds show distinct environmental conditions, including water temperature, water chemistry, substrate, and flow/sediment regimes[@altermattDiversityRiverineMetacommunities2013; @bendaNetworkDynamicsHypothesis2004; @mooreEmergentStabilityLarge2015; @nakamuraDisturbanceRegimesStream2000; @petersonModellingDendriticEcological2013; @teruiMetapopulationStabilityBranching2018]. Instead, it may be more relevant to severely altered landscapes where human disturbance compromises the environmental distinctiveness of branches through, for example, flow regulations by dams[@poffHomogenizationRegionalRiver2007]. Hence, our theory has important implications for riverine biodiversity conservation by pointing the crucial role of habitat diversity produced by branching structure.

@@

Despite the substantial difference in fish fauna between the study regions, the estimated Bayes factor in favor of the global model was `r round(fit$gamma$bayes_factor, 1)` for $\gamma$ diversity; this result strongly supports the consistent effects of ecosystem size and complexity on $\gamma$ diversity across the two biogeographic regions. Further, we found patterns consistent with our theoretical predictions. The estimated $\gamma$ diversity increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) across watersheds. These effects remained significant even after controlling for the potential influences of other environmental factors (precipitation, temperature, and land use). Overall, our statistical analysis provides strong empirical evidence that ecosystem size and complexity jointly, but independently, scale riverine biodiversity at the metacommunity level.

Similarly, we found weak to moderate supports for the global models of $\alpha$ and $\beta$ diversity (Bayes factor: `r round(fit$alpha$bayes_factor, 1)` and `r round(fit$beta$bayes_factor, 1)`, respectively). In both regions, $\beta$ diversity responded significantly to ecosystem size and complexity while $\alpha$ diversity showing a vague response to these ecosystem properties. In our simulations, this pattern has emerged under the scenarios with dispersal limitation, which elegantly match the previous observations of stream fish movement. Direct (mark-recapture) and indirect observations (e.g., genotyping) recurrently revealed restricted movement of stream fish, typically limited to several tens to hundreds of meters in distance[@comteFishDispersalFlowing2018; @radingerPatternsPredictorsFish2014; @rodriguezRestrictedMovementStream2002]. The reciprocal agreement of theoretical and empirical patterns provides indirect but convincing evidence that dispersal limitation plays a key role in driving the associations between $\gamma$ diversity and ecosystem properties in rivers.

The consistent effect of branching probability on $\gamma$ diversity across the study regions is noteworthy because the watersheds in the Midwest are severely altered by agriculture (median % agricultural land use: `r round(median(wsd_subset[[2]]$frac_agri)*100)`% for Midwest and `r round(median(wsd_subset[[1]]$frac_agri)*100)`% for Hokkaido). If the intensive land use by humans causes significant homogenization of in-stream habitat conditions among tributaries, theory predicts weakened effects of branching probability on $\gamma$ diversity. However, $\gamma$ diversity increased significantly with increasing branching probability in this highly modified landscape, suggesting that tributaries still sustain unique environmental conditions to support high spatial trunover of species composition. In support of this interpretation, $\beta$ diversity increased with increasing branching probability in both regions (**Figure 4**). It is conceivable that local geological and geomorphological differences, such as slope, aspect, and soil porosity, still persist in human-dominated landscapes to maintain diversity of in-stream processes among tributaties[@bendaNetworkDynamicsHypothesis2004]. Although our analysis is correlative and cannot provide conclusive evidence, the finding is encouraging because branching complexity of river networks may serve as a natural defense system to human-induced environmental changes.

@@

The emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation[@poffHomogenizationRegionalRiver2007], habitat fragmentation[@fukushimaModellingEffectsDams2007; @moritaEffectsHabitatFragmentation2002], and stream burial, may compromise or restrict access to the diverse habitats that complex branching networks may provide. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial arrangement of conservation efforts. A growing body of evidence suggests the effectiveness of riverine reserves and local restoration, and how to arrange a spatial network of conservation sites in rivers is an area of active research[@koningNetworkGrassrootsReserves2020; @stollSmallImpoverishedRegional2013; @sundermannRiverRestorationSuccess2011]. The results of this study clearly indicate that riverine conservation networks should be designed to involve diverse tributaries, as inferred from a successful reserve network of the Salween basin in Thailand[@koningNetworkGrassrootsReserves2020].

@@

where $\mu_{z,k}$ is the environmental value at longitudinal position $k$ ($k$ is the network distance from the outlet patch; $k = 1$ at the outlet), $\rho$ is the strength of spatial autocorrelation and $\epsilon_k$ is the local environmental noise that follows a normal distribution with a mean of zero and SD of $\sigma_l$. The parameter $\rho$ can take values of $0-1$ with larger values indicating greater spatial autocorrelation. In this study, we set $\rho = 1$ to mimic strong spatial autocorrelation in rinverine environments. At confluences, we took a weighted mean of environmental values given the relative size of upstream contributing area of joining tributaries:

@@

where $r_{ix}(t)$ is the reproductive rate of species $i$ given the environmental condition at patch $x$ and time $t$, $r_{0,i}$ the maximum reproductive rate of species $i$, $K_{x}$ the carrying capacity at patch $x$, $\alpha_{ij}$ the interaction coefficient between species $i$ and $j$, and $S$ the number of species in a metacommunity. The parameter $\alpha_{ij}$ is expressed as the strength of interspecific competition relative to that of intraspecific competition such that interspecific competition is greater than intraspecific competition if $\alpha_{ij}$ > 1 (intraspecific competition coefficient $\alpha_{ii} = 1$). The density-independent reproductive rate $r_{ix}(t)$ is affected by abiotic environments (non-consumable) and determined by the following Gaussian function:

@@

We compiled fish data at two biogeographic regions: Hokkaido island, Japan and Midwest, US. For Hokkaido, we used data from the Hokkaido Freshwater Fish Database HFish, monitoring data at protected watersheds, and primary data collected from literature. Collectively, these fish data encompass more than 7000 sites and covers the entire Hokkaido island. For the Midwest, we assembled fish community data collected by XXX. We detailed procedures for initial data screening in SI.

We further selected data based on the following criteria. First, we selected sampling sites with reliable coordinates. Sampling sites were treated as the same site if coordinates are identical at 3 decimal degrees. Second, for sites with multiple site visits (i.e., temporal replicates), we used the latest data at each sampling site to minimize variation in sampling efforts among sites. Lastly, we selected metacommunities (i.e., watersheds) that contain more than 10 sampling sites to reduce statistical uncertainties in estimating $\gamma$ diversity. We considered watersheds as separated from others if the river network flows into one of the following: the ocean, a large lake ($\ge$ 10 km² in the areal area) or a large river that may represent lentic habitats ($\ge$ 5000 km² in watershed area).

@@

$$
\mu_i = \xi_0 + \xi_1~log_{10}~wa_i + \xi_2~log_{10}~p_{b,i} + \xi_3~region_i + \xi_4~log_{10}~WA_i \cdot region_i + \xi_5~log_{10}~P_{b,i} \cdot region_i + \sum_k \xi_k x_{k,i}
$$
We compared the performance of these competing models using the Bayes factor[@wagenmakersPracticalSolutionPervasive2007], which gives the relative strength of evidence in favor of the global model (M0) over the region-specific model (M1). We used Bayesian Information Criteria (BIC) of the linear models to approximate the Bayes factor[@wagenmakersPracticalSolutionPervasive2007]:

>>>>

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of Minnesota]
- Taku Kadoya^[National Institute for Environmental Studies]
- Yusuke Miyazaki^[Shiraume Gakuen University]
output:
  word_document: default
  html_notebook: default
abstract:
  The species-area relationship (the SAR), a tendency in which species richness increases with increasing area, is one of the most robust empirical generalizations in ecology and has provided the foundation for area-based conservation strategies. However, many ecosystems possess complex spatial structure that cannot be repserented by area (the scale-invariant complexity) and its role in scaling biodiversity is largely unknown. Here, we hypthesized that ecosystem size and complexity jointly regulate biodiversity in rivers, a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns. Ecosystem size should dictate metacommunity size and habitat diversity that regulate watershed-scale species richness. Meanwhile, ecosystem complexity, which we defined as the probability of branching per unit river distance, should increase watershed-scale species richness by increasing habitat heterogeneity. We developed theoretical models simulating metacommunity dynamics in branching river networks and yielded results consistent with our hypothesis. Ecosystem size and complexity both increased watershed-scale species richness ($\gamma$ diversity), and the importance of spatial turnover in species composition ($\beta$ diversity) was apparent when simulated species travel short distances. Our cross-watershed comparisons across two biogeographic regions (Hokkaido, Japan and Midwest, US) provided empirical supports for the theoretical predictions. As predicted, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial turnover in species composition over the branching network. There is now a clear need to consider the dimension of ecosystem complexity in biodiversity conservation. The dual dimensional approach - size and complexity - should broaden viable options for spatial planning of protected areas or restoration sites, thereby helping reconcile the conflicts between biodiversity conservation and societal demands.

bibliography: reference.bib
csl: nature.csl
---

@@

# Maintext

@@

Influences of ecosystem size and complexity differed significantly in their dependence on landscape scenarios (**cite SI Figure**). Ecosystem size had positive effects on $\gamma$ diversity regardless of landscape scenarios, although the slopes were steeper with greater environmental variation (higher $\sigma_h$ and/or $\sigma_l$) (**cite SI Figure**). This result is attributable to the fact that larger ecosystems can hold more individuals in a metacommunity[@hubbellUnifiedNeutralTheory2001]. In contrast, we observed limited or no influences of branching complexity when local environmental noise was equal to or exceeded environmental variation at headwaters ($\sigma_l \ge \sigma_h$). Under this scenario, having more branches may not contribute to the overall habitat diversity in the ecosystem because local environmental noise masks environmental differences between branches (**cite SI Figure**). Therefore, this landscape scenario decouples the intimate relationship between branching structure and habitat diversity, thereby eliminating the positive effect of branching complexity on $\gamma$ diversity. This theoretical prediction may not apply to pristine or semi-natural river networks where individual streams with diverse geological and/or climatic backgrounds show distinct environmental conditions, including water temperature, water chemistry, substrate, and flow/sediment regimes[@altermattDiversityRiverineMetacommunities2013; @bendaNetworkDynamicsHypothesis2004; @mooreEmergentStabilityLarge2015; @nakamuraDisturbanceRegimesStream2000; @petersonModellingDendriticEcological2013; @teruiMetapopulationStabilityBranching2018]. Instead, it may be more relevant to severely altered landscapes where human disturbance compromises the environmental distinctiveness of branches through, for example, flow regulations by dams[@poffHomogenizationRegionalRiver2007]. Hence, our theory has important implications for riverine biodiversity conservation by pointing the crucial role of habitat diversity produced by branching structure.

@@

Despite the substantial difference in fish fauna between the study regions, the estimated Bayes factor in favor of the global model was `r round(fit$gamma$bayes_factor, 1)` for $\gamma$ diversity; this result strongly supports the consistent effects of ecosystem size and complexity on $\gamma$ diversity across the two biogeographic regions. We found patterns consistent with our theoretical predictions (**Table 1 and Figure 4**). The estimated $\gamma$ diversity increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) across watersheds. These effects remained significant even after controlling for the potential influences of other environmental factors (precipitation, temperature, and land use) (**Table 1**). Overall, our statistical analysis provides strong empirical evidence that ecosystem size and complexity jointly, but independently, scale riverine biodiversity at the metacommunity level.

Similarly, we found weak to moderate supports for the global models of $\alpha$ and $\beta$ diversity (Bayes factor: `r round(fit$alpha$bayes_factor, 1)` and `r round(fit$beta$bayes_factor, 1)`, respectively). In both regions, $\beta$ diversity responded significantly to ecosystem size and complexity while $\alpha$ diversity showing a vague response to these ecosystem properties (**Table 1 and Figure 4**). In our simulations, this pattern has emerged under the scenarios with dispersal limitation, which elegantly match the previous observations of stream fish movement. Direct (mark-recapture) and indirect observations (e.g., genotyping) recurrently revealed restricted movement of stream fish, typically limited to several tens to hundreds of meters in distance[@comteFishDispersalFlowing2018; @radingerPatternsPredictorsFish2014; @rodriguezRestrictedMovementStream2002]. The reciprocal agreement of theoretical and empirical patterns provides indirect but convincing evidence that dispersal limitation plays a key role in driving the associations between $\gamma$ diversity and ecosystem properties in rivers.

The consistent effect of branching probability on $\gamma$ diversity across the study regions is noteworthy because the watersheds in the Midwest are severely altered by agriculture (median % agricultural land use: `r round(median(wsd_subset[[2]]$frac_agri)*100)`% for Midwest and `r round(median(wsd_subset[[1]]$frac_agri)*100)`% for Hokkaido). If the intensive land use by humans causes significant homogenization of in-stream habitat conditions among tributaries, theory predicts weakened effects of branching probability on $\gamma$ diversity. However, $\gamma$ diversity increased significantly with increasing branching probability in this highly modified landscape, suggesting that tributaries still sustain unique environmental conditions to support high spatial turnover of species composition. In support of this interpretation, $\beta$ diversity increased with increasing branching probability in both regions (**Figure 4**). It is conceivable that local geological and geomorphological differences, such as slope, aspect, and soil porosity, still persist in human-dominated landscapes to maintain diversity of in-stream processes among tributaties[@bendaNetworkDynamicsHypothesis2004]. Although our analysis is correlative and cannot provide conclusive evidence, the finding is encouraging because branching complexity of river networks may serve as a natural defense system to human-induced environmental changes.

@@

The emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation[@poffHomogenizationRegionalRiver2007], habitat fragmentation[@fukushimaModellingEffectsDams2007; @moritaEffectsHabitatFragmentation2002], and stream burial, may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and how to design a spatial network of conservation sites in rivers is an area of active research[@koningNetworkGrassrootsReserves2020; @stollSmallImpoverishedRegional2013; @sundermannRiverRestorationSuccess2011]. While large reserves or restoration sites are undoubtedly important[@koningNetworkGrassrootsReserves2020], our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when multiple ecologically-distinct tributaties are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin[@koningNetworkGrassrootsReserves2020]. Our findings, therefore, offer the ecological foundation for the network-based conservation strategies in rivers, broadening viable options towards successful conservation.

@@

where $\mu_{z,k}$ is the environmental value at longitudinal position $k$ ($k$ is the network distance from the outlet patch; $k = 1$ at the outlet), $\rho$ is the strength of spatial autocorrelation and $\epsilon_k$ is the local environmental noise that follows a normal distribution with a mean of zero and SD of $\sigma_l$. The parameter $\rho$ can take values of $0-1$ with larger values indicating greater spatial autocorrelation. In this study, we set $\rho = 1$ to mimic strong spatial autocorrelation in riverine environments. At confluences, we took a weighted mean of environmental values given the relative size of upstream contributing area of joining tributaries:

@@

where $r_{ix}(t)$ is the reproductive rate of species $i$ given the environmental condition at patch $x$ and time $t$, $r_{0,i}$ the maximum reproductive rate of species $i$, $K_{x}$ the carrying capacity at patch $x$, $\alpha_{ij}$ the interaction coefficient between species $i$ and $j$, and $S$ the number of species in a metacommunity. $K_x$ was expressed as a function of the number of upstream contributing patches $A_x$ ($K_x = 100A_x$) to represent increasing habitat size with increasing upstream watershed area[@altermattRiverNetworkProperties2013; @teruiThreeEcologicalFactors2016]. The parameter $\alpha_{ij}$ is expressed as the strength of interspecific competition relative to that of intraspecific competition such that interspecific competition is greater than intraspecific competition if $\alpha_{ij}$ > 1 (intraspecific competition coefficient $\alpha_{ii} = 1$). $\alpha_{ij}$ was drawn randomly from a uniform distribution as $\alpha_{ij} \sim Unif(0, \alpha_{max})$. The density-independent reproductive rate $r_{ix}(t)$ is affected by abiotic environments (non-consumable) and determined by the following Gaussian function:

@@

We compiled fish data at two biogeographic regions: Hokkaido island, Japan and Midwest, US. For Hokkaido, we used data from the Hokkaido Freshwater Fish Database HFish, monitoring data at protected watersheds, and primary data collected from literature, covering the entire Hokkaido island. For the Midwest, we assembled fish community data collected by Iowa Department of Natural Resources, Illinois Environmental Protection Agency and Illinois Department of Natural Resources, Minnesota Pollution Control Agency, and Wisconsin Department of Natural Resources. We detailed procedures for initial data selection in SI.

We selected data based on the following criteria. First, we selected sampling sites with reliable coordinates. Sampling sites were treated as the same site if coordinates are identical at 3 decimal degrees. Second, for sites with multiple site visits (i.e., temporal replicates), we used the latest data at each sampling site to minimize variation in sampling efforts among sites. Lastly, we selected metacommunities (i.e., watersheds) that contain more than 10 sampling sites to reduce statistical uncertainties in estimating $\gamma$ diversity. We considered watersheds as separated from others if the river network flows into one of the following: the ocean, a large lake ($\ge$ 10 km² in the areal area) or a large river that may represent lentic habitats ($\ge$ 5000 km² in watershed area).

@@

$$
\mu_i = \xi_0 + \xi_1~log_{10}~WA_i + \xi_2~log_{10}~P_{b,i} + \xi_3~region_i + \xi_4~log_{10}~WA_i \cdot region_i + \xi_5~log_{10}~P_{b,i} \cdot region_i + \sum_k \xi_k x_{k,i}
$$
We compared the performance of these competing models using the Bayes factor[@wagenmakersPracticalSolutionPervasive2007], which gives the relative strength of evidence in favor of the global model (M0) over the region-specific model (M1). We used Bayesian Information Criteria (BIC) of the linear models to approximate the Bayes factor[@wagenmakersPracticalSolutionPervasive2007]:
